---
title: "crudr"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{crudr}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  markdown: 
    wrap: 72
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(crudr)
```

## 

# Overview

This package is a CRUD app. The idea behind crudr is to make it easy for
you or your organization to centrally maintain a single source of data
while at the same time allowing distributed control of the data for
updates.


## A Common Problem

Each company I have worked for has different departments with different
spreadsheets that track the same information. For instance, let's say
you have a bunch of store locations in your company. Ideally, someone in
your company has all the updated information about those locations, e.g.
the store's mailing address, phone number, email, manager, manager's
email, etc. In most companies, that information actually resides in a
lot of different places and it almost always has errors--managers move
around, store leases expire, and outdated spreadsheets are all over the
company. Further, Joe might be in charge of the locations, but Sally
knows the Manager info, etc. You need that data to be correct, joined,
posted, and available for everyone, but that's hard to do. This package
helps to solve this distributed updating and distributed use problem.

## Goals for crudr

The goals with crudr are to: 

1. Present data from a relational database
through a Shiny app for viewing and management 
2. Make it easy to control which admin users can update columns in a DB table 
3. Keep a record of all changes made to the database table in a corresponding
change log table 
4. Allow more than one simultaneous editor without crashes. 
5. Be database agnostic.


### Other Notes

-   Crudr works by synching tables between the Admin User machines, the
    Host Server (my desktop), and the Database
-   The code is built with modules so that you can manage and present
    several tables in the same app
-   Databases that as been tested are SQLite, Postgres, and Snowflake
-   The code is in development, so names and parameters aren't
    yet decided, but they're pretty good. 
-   The database checks periodically of others made changes to the data; if so, the UI and server refresh. 
-   I would love if someone at RStudio could improve the code or just
    take the code and make it their own. It'd be great if people could use the 
    app broadly. I don't know how to make it really robust; I'm a novice. 

##

# Getting Started

To install crudr, run:

```{r install-crudr, eval = F}
remotes::install_github('eauleaf/crudr')
```

The Example is built from these packages, so you might want to install:

Note: pool needs the most updated version on CRAN. 

```{r packages, eval = F}
install.packages(c(
  'shiny',
  'shinydashboard',
  'tidyverse',
  'lubridate',
  'DT',
  'pool'
))
```

## 2 crudr functions

There are really only 2 crudr functions you need to know:
**`cdr_make_db_tbls()`** & **`cdr_manage_db_tbls()`**

### cdr_make_db_tbls() 
This function writes two tables into your
database. You hand this function a tibble and a pool connection, and
it'll write that table and a corresponding change tracking
table in your database. If you already have the table, tell `cdr_make_db_tbls()` the
name of that database table and the function will then only create the
change tracking table. If you already have both tables, the function
doesn't do anything except write some notes out to your R console saying
it's not going to mess with tables that already exist. Note: For crudr
to work, each row of the table you write must have a unique ID that's
recorded as a character. If you don't have a unique ID key field in your
table, `cdr_make_db_tbls()` will automatically make one for you named "UID".


##### Example for cdr_make_db_tbls():
For example, if you want to
create a database on your machine called 'test.db' and want to put in a
table called 'iris', you'd write:

```{r sqlite-db, eval = F}
con <- pool::dbPool(DBI::dbConnect(RSQLite::SQLite(), 'test.db'))
```

Then you'd call the function with the above connection and a table to
write.

```{r create-tbls, eval = F}
cdr_make_db_tbls(con, iris)
```

If you already had a Unique ID key field in your table, just tell the
function the name of your field with `cdr_make_db_tbls(con, iris,
key_field = 'name_of_your_unique_id_field')`.

Okay, the function should have written in your database a table called 'iris' 
and a second table called 'iris_DELTAS'. The table names are case sensitive. You can see the tables that the function wrote by running the following:

```{r eval = F}
paste(pool::dbListTables(con), collapse = ', ')
```

You should see the printout: "iris" and "iris_DELTAS". These two tables are your original table,
which is the the primary table, and a secondary table named like your first, but
with '\*\_DELTAS'. Your primary table will now have a unique ID field if it
didn't already, and the secondary deltas table is a change log. Any change you
make to the primary table through the admin user interface, which we'll soon do,
gets recorded in the deltas table.

As a side note, if you want to remove the tables later, you can that with:

```{r eval = F}
pool::dbRemoveTable(con,'iris')
pool::dbRemoveTable(con,'iris_DELTAS')
```

Or, just delete the database 'test.db'.

### cdr_manage_db_tbls() 

This function is the workhorse of the package. If you remember one function, this is the one. 
The function only has a handful of arguments. You have to tell this function: 

1. What the name of your table is. In this case it's, *db_tbl_name = 'iris'*.
1. The field with your Unique ID. In this case it's *key_col = 'UID'*.
1. What the database connection is. In this case it's, *conn_pool = con*.

That's about it. There are other arguments you can change if you want to control 
who can change things in the primary (admin) table. 

`cdr_manage_db_tbls()` returns 1 table directly and 2 tables invisibly.

The direct table output in the above example is: `iris_r_tbl`

When it comes out of Shiny, you call the table by acting like it's a function, so 
you say `iris_r_tbl()`. This table is yours; you can do whatever you want with it. 
You can send it out in any format you like, change the column names, filter it, 
join it to something else... Whatever you like. It's an output. 

The 2 tables that `cdr_manage_db_tbls()` returns invisibly are 'db_tbl' and 'chg_log_tbl'. These
are not really changeable like the direct output table. Rather, these tables are direct representations of what's in your database. 

### db_tbl 
This invisible table is your administrator table. It's the database table you wrote into the 
database, so in our case, it should be the iris data. Whoever maintains your 
dataset gets access to db_tbl, and can make changes to the data held in the database. 
You collect and present this table in Shiny by calling it with function:

**crudr::cdr_admin_tbl_ui('iris', 'db_tbl')**.

### chg_log_tbl 
This invisible table keeps track of what changes you made to 'db_tbl'. You never need to do 
much with this table. It's just there to record which administrator did what to the 
database table, and when they made the change. It should be empty right now, but 
let's change that. You collect and present this table in Shiny by calling it with function:

**crudr::cdr_admin_tbl_ui('iris', 'chg_log_tbl')**.


##### One more function to note: 

**crudr::cdr_admin_key_ui()**

`crudr::cdr_admin_key_ui()` writes the html for admin changes to add or 
delete rows. Just pass it the name of your database table in quotes, like 'iris'. 



Okay. We're done. Let's launch an app.


##

# Launch a Tiny Shiny App

To manage the database tables, now let's launch a bare-bones shiny app. We'll 
only need the functions described above from crudr. The other stuff comes from 
packages *shiny*, *shinydashboard*, and *DT*. 


Here's the code for your app. Copy it into your RStudio console and run it. 


## CRUDR Example 
```{r tiny_crudr_app, eval = F}

con <- pool::dbPool(DBI::dbConnect(RSQLite::SQLite(), 'test.db'))

crudr::cdr_make_db_tbls(con, iris)

header <- shinydashboard::dashboardHeader(title = 'Tiny CRUDR Example')

sidebar <- shinydashboard::dashboardSidebar(
  shinydashboard::sidebarMenu(
    id = 'tabs',
    shinydashboard::menuItem(
      text = "Iris Data",
      startExpanded = TRUE,
      shinydashboard::menuSubItem("Administrator Table", tabName = "datr_editable", icon = shiny::icon('edit')),
      shinydashboard::menuSubItem("Change Log", tabName = "datr_change_log"),
      shinydashboard::menuSubItem("Iris End User View", tabName = "datr_end_usr")
    )))

body <- shinydashboard::dashboardBody(
  shinydashboard::tabItems(
    shinydashboard::tabItem(tabName = "datr_editable", htmltools::div(crudr::cdr_admin_key_ui('iris'), crudr::cdr_admin_tbl_ui('iris', 'db_tbl'))),
    shinydashboard::tabItem(tabName = "datr_change_log", crudr::cdr_admin_tbl_ui('iris', 'chg_log_tbl')),
    shinydashboard::tabItem(tabName = "datr_end_usr", DT::DTOutput('iris_db_data'))
  ))

ui <- shinydashboard::dashboardPage(header, sidebar, body)



server <- function(input, output, session){


  iris_r_tbl <- crudr::cdr_manage_db_tbls(
   db_tbl_name = 'iris',
    key_col = 'UID',
    conn_pool = con,
    session = session,
    add_row_permission = T,
    del_row_permission = T,
    cell_edit_permission = T,
    lock_fields = c()
  )

  output$iris_db_data <- DT::renderDT(
    DT::formatDate(table   = DT::datatable(iris_r_tbl()),
                   columns = c('WHEN_EDITED_LAST'),
                   method  = 'toLocaleString')
  )



}


shiny::shinyApp(ui, server)

```

# Make Changes

In the app above, you're the admin. So, go ahead and make changes in this admin table. 
Double-click on a cell and change the data. Your app then writes that data to the 
database, updates the change log table, write the change log data to the database, and 
then waits for your next change. 

That's pretty much it. If you want to set permissions, you can do so in your 
shiny server with something like:

```{r user_control, eval = F}

user <- ifelse(is.null(session$user), Sys.info()[['user']], session$user)

if (user == 'my_name'){
  edit_permission = T
} else {
  edit_permission = F
}

```

Then pass edit_permission into the `cdr_manage_db_tbls()` permission argument(s). That
makes the admin table just another table for anyone but you.


## Other stuff


For a Postgres connection, use `RPostgreSQL::PostgreSQL()` as the driver, since 
that's what I tested on. Example Postgres connection below. 

```{r postgres_connection, eval = F}
con = pool::dbPool(
           DBI::dbConnect( RPostgreSQL::PostgreSQL(),
                           dbname = "test",
                           host = "localhost",
                           port = "5432",
                           user = Sys.getenv('dsn_uid'),
                           password = Sys.getenv('dsn_pwd')))
```

To close your database connection, run: `pool::poolClose(con)`


There are quite a few things I need to do to make this better. Maybe I'll get to these
sometime. 

1. Support for more databases, like sqlserver, oracle, etc. (I don't think 
it'd be that hard, I just need a test connection.)
2. At some point I should create a function where you give a date and the function
rebuilds the database with the data as it was at that particular point in time. 
3. Figure out how to push data to the database asynchronously so the user never 
has to wait. 
4. Write automatic tests for the functions. 
5. Figure out package exports so I'm not exporting everything. 


